Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-14 paper.qmd
Updated equations to inline
Updated the remaining equations

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = r_2/r_1$ (@fig-NLDR-two-curvy). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1$. <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

The value for $b_2$ is computed by fixing $b_1$. Considering the upper bound of the first NLDR component, $a_1 > (1+2q)/(b_1 -1)$. Similarly, for the second NLDR component, $a_2 > (r_2 + q(1 + r_2))/(b_2 - 1)$. Since $a_2 = \sqrt(3)a_1/2$ for regular hexagons, $a_1 > \frac{2[r_2 + q(1 + r_2)]}{\sqrt{3}(b_2 - 1)}$. This is a linear optimization problem. Therefore, the optimal solution must occur on a vertex. Therefore, $b_2 = \Big\lceil1 +\frac{2[r_2 + q(1 + r_2)](b_1 - 1)}{\sqrt{3}(1 + 2q)}\Big\rceil$.

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{\frac{n}{r_2}}$ are allowed. @fig-param-two-curvy (a) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

The area of a hexagon is defined as $A = \frac{3\sqrt{3}}{2}l^2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\frac{\sum^{h}_{i=1}n_h}{A}$ and the proportion is $\frac{\sum^{h}_{i=1}n_h}{A \times b}$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.

>>>>

Because we are working with distances between points, starting with data having a standard scale, e.g. [0, 1], is recommended. The default should take the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$ into account. When $k=2$, as in hexagon binning, the default range is $[0, y_{i,\text{max}}], i=1,2$, where $y_{1,\text{max}}=1$ and $y_{2,\text{max}} = r_2/r_1$ (@fig-NLDR-two-curvy). If the NLDR aspect ratio is ignored then set $y_ {2,\text{max}} = 1.$ <!-- \times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height ($a_1$) and width ($a_2$) of a regular hexagon.) The scaling of data should take the size of the hexagons into account, but choice of number of bins should. -->

@@

The value for $b_2$ is computed by fixing $b_1$. Considering the upper bound of the first NLDR component, $a_1 > (1+2q)/(b_1 -1)$. Similarly, for the second NLDR component,

$$
a_2 > \frac{r_2 + q(1 + r_2)}{(b_2 - 1)}.
$$

Since $a_2 = \sqrt{3}a_1/2$ for regular hexagons,

$$
a_1 > \frac{2[r_2 + q(1 + r_2)]}{\sqrt{3}(b_2 - 1)}.
$$

This is a linear optimization problem. Therefore, the optimal solution must occur on a vertex. Therefore,

$$
b_2 = \Big\lceil1 +\frac{2[r_2 + q(1 + r_2)](b_1 - 1)}{\sqrt{3}(1 + 2q)}\Big\rceil.
$$

@@

The default number of bins $b=b_1\times b_2$ is computed based on the sample size, by setting $b_1=n^{1/3}$, consistent with the Diaconis-Freedman rule [@freedman1981]. The value of $b_2$ is determined analytically by $b_1, q, r_2$. Values of $b_1$ between $2$ and $b_1 = \sqrt{n/r_2}$ are allowed. @fig-param-two-curvy (a) shows the effect of different choices of $b_1$ on the MSE of the fitted model.

@@

The area of a hexagon is defined as $A = 3\sqrt{3}l^2/2$ where $l$ is the side length of the hexagon. If we know $a_1$ and $a_2$, $l$ can be computed (see appendix). The density of a hexagon grid is calculated as $\sum^{h}_{i=1}n_h/A$ and the proportion is $\sum^{h}_{i=1}n_h/Ab$. The baseline proportion is the proportion density at the smallest possible value of $a_1$. The relative proportion density is the ratio of the observed proportion density to the baseline proportion density.
